---
title: "Practical Machine Learning Project Writeup"
author: "Madhu Balaji"
date: "August 23, 2015"
output: 
  html_document: 
    keep_md: yes
    smart: no
    toc: yes
---

<b>The goal of the project:</b>

To predict the manner in which they did the exercise.

Assumption:- Data file is downloaded and stored into a folder "data"
</br>
<b>Background:</b></br>
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

</br>
<b>Data Processing:</b></br>

Loading the package:</br>

```{r}

library(AppliedPredictiveModeling)
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
set.seed(4680)
```

Getting the data: </br>

The Training Data

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
The Testing Data

```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Two options to use the data:
1. Access the URL and load in to the memory

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

2. Download file and store on your harddisk

```{r}
##training <- read.csv("data/pml-training.csv", header=TRUE, as.is = TRUE, na.strings = c('NA','','#DIV/0!'))
##testing <- read.csv("data/pml-testing.csv", header=TRUE, as.is = TRUE, na.strings = c('NA','','#DIV/0!'))
```

<b>Partition the data</b></br>
Partitioning data set into 60 - 40 % for training and testing respectively

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTrain <- training[inTrain, ]
myTest <- training[-inTrain, ]
dim(myTrain) 
dim(myTest)
```
<b>Cleaning the data</b></br>
Transformations used to clean the data:

Transformation 1: Cleaning NearZeroVariance Variables Run this code to view possible NZV Variables

```{r}
myData1 <- nearZeroVar(myTrain, saveMetrics=TRUE)
```
Create another subset without NZV variables:

```{r}
myNZV <- names(myTrain) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
                                      "kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
                                      "max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
                                      "var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
                                      "stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
                                      "kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
                                      "max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
                                      "kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
                                      "skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
                                      "amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
                                      "skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
                                      "max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
                                      "amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
                                      "avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
                                      "stddev_yaw_forearm", "var_yaw_forearm")
myTrain <- myTrain[!myNZV]

dim(myTrain)
```

Transformation 2: Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms

```{r}
myTrain <- myTrain[c(-1)]
```
Transformation 3: Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA’s I’m going to leave them out

```{r}
trainingVal <- myTrain #creating another subset to iterate in loop
for(i in 1:length(myTrain)) { #for every column in the training dataset
        if( sum( is.na( myTrain[, i] ) ) /nrow(myTrain) >= .6 ) { #if n?? NAs > 60% of total observations
                for(j in 1:length(trainingVal)) {
                        if( length( grep(names(myTrain[i]), names(trainingVal)[j]) ) ==1)  { #if the columns are the same:
                                trainingVal <- trainingVal[ , -j] #Remove that column
                        }   
                } 
        }
}
dim(trainingVal)
```
```{r}
myTrain <- trainingVal
rm(trainingVal)
```

Now, let's transform testing data set

```{r}
clean1 <- colnames(myTrain)
clean2 <- colnames(myTrain[, -58]) #already with classe column removed
myTest <- myTest[clean1]
testing <- testing[clean2]

dim(myTest)

dim(testing)
```
Just to ensure proper functioning of Decision Trees with the Test data set, we need to coerce the data into the same type:

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTrain)) {
                if( length( grep(names(myTrain[i]), names(testing)[j]) ) ==1)  {
                        class(testing[j]) <- class(myTrain[i])
                }      
        }      
}

testing <- rbind(myTrain[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```
Using algorithms for prediction: </br>
<b>Decision Tree</b>

```{r}
mFitDT <- rpart(classe ~ ., data=myTrain, method="class")

fancyRpartPlot(mFitDT)
```

Predicting:

```{r}
predictDT <- predict(mFitDT, myTest, type = "class")
```

Using confusion Matrix to test the results:
```{r}
confusionMatrix(predictDT, myTest$classe)
```

```{r}
##Confusion Matrix and Statistics

##          Reference
##Prediction    A    B    C    D    E
##         A 2147   66    6    4    0
##         B   62 1252   88   69    0
##         C   23  188 1244  137    1
##         D    0   12   21  872   92
##         E    0    0    9  204 1349

##Overall Statistics
                                          
##               Accuracy : 0.8748          
##                 95% CI : (0.8673, 0.8821)
##    No Information Rate : 0.2845          
##    P-Value [Acc > NIR] : < 2.2e-16       
                                          
##                  Kappa : 0.8416          
## Mcnemar's Test P-Value : NA              

##Statistics by Class:

##                     Class: A Class: B Class: C Class: D Class: E
##Sensitivity            0.9619   0.8248   0.9094   0.6781   0.9355
##Specificity            0.9865   0.9654   0.9461   0.9809   0.9667
##Pos Pred Value         0.9658   0.8511   0.7809   0.8746   0.8636
##Neg Pred Value         0.9849   0.9583   0.9802   0.9396   0.9852
##Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
##Detection Rate         0.2736   0.1596   0.1586   0.1111   0.1719
##Detection Prevalence   0.2833   0.1875   0.2030   0.1271   0.1991
##Balanced Accuracy      0.9742   0.8951   0.9277   0.8295   0.9511

```

<b>Random Forests</b>

```{r}
mFitRF <- randomForest(classe ~. , data=myTrain)
```

Predicting in-sample error:

```{r}
predictRF <- predict(mFitRF, myTest, type = "class")
```

Using confusion Matrix to test the results:

```{r}
confusionMatrix(predictRF, myTest$classe)
```
```{r}
##Confusion Matrix and Statistics

##          Reference
##Prediction    A    B    C    D    E
##         A 2232    0    0    0    0
##         B    0 1518    3    0    0
##         C    0    0 1364    2    0
##         D    0    0    1 1282    0
##         E    0    0    0    2 1442

##Overall Statistics
                                         
##               Accuracy : 0.999          
##                 95% CI : (0.998, 0.9996)
##    No Information Rate : 0.2845         
##    P-Value [Acc > NIR] : < 2.2e-16      
                                         
##                  Kappa : 0.9987         
## Mcnemar's Test P-Value : NA             

##Statistics by Class:

##                     Class: A Class: B Class: C Class: D Class: E
##Sensitivity            1.0000   1.0000   0.9971   0.9969   1.0000
##Specificity            1.0000   0.9995   0.9997   0.9998   0.9997
##Pos Pred Value         1.0000   0.9980   0.9985   0.9992   0.9986
##Neg Pred Value         1.0000   1.0000   0.9994   0.9994   1.0000
##Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
##Detection Rate         0.2845   0.1935   0.1738   0.1634   0.1838
##Detection Prevalence   0.2845   0.1939   0.1741   0.1635   0.1840
##Balanced Accuracy      1.0000   0.9998   0.9984   0.9984   0.9998
```

<b>Prediction Analysis - Results</b>

<b>Random Forests yielded better Results. It was slow but accurate with accuracy of 99.9% <b>

Final in-smaple prediction

```{r}


predictionFinal <- predict(mFitRF, testing, type = "class")
```